Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment

### 2017-04-14 Latex/article.tex
More Bildhauer comments, PhilStat moved to Sec 3.

Before moving on to the description of the phenomenon and the corpus study in Sections~\ref{sec:descriptive} and \ref{sec:analyses}, I comment on the statistical methods used in the corpus study in Section~\ref{sec:rightstatistics} because the discussion is closely related to basic questions of the cognitive approach in corpus linguistics.

In this section, I motivate the choice of statistical models which I use later in Section~\ref{sec:corpusstudies}.
Readers might think that the method used here for modeling grammatical alternations -- namely (Hierarchical) Logistic Regression\slash Generalised Linear (Mixed) Models or GLM(M)s -- does not require much motivation.
After all, GLMMs have been established as the major tool in the analysis of alternation phenomena.
All studies mentioned at the outset of Section~\ref{sec:cogocl} use some form of regression\slash GLM(M).
Over the past few years, however, modified or alternative methods have been proposed.
From among these methods, I just make a few remarks on Bayesian estimation (see \citealp{GelmanEa2014}), as it was proposed in \cite{Levshina2016,Divjak2016a}, for example.
Conceptually, I see three points of discussion that should be kept apart.
First, Bayesian methods are sometimes touted as superior tools for scientific inference compared to frequentist methods.
Second, it has been proposed that the Bayesian interpretation of probability is more cognitively adequate for the modeling of linguistic data.
Third, and very specific to this paper, given established methods in the modeling of alternation and variation, it has to be decided whether so-called Bayesian methods lead to substantially different results.

This concludes the general introduction.
In the remainder of the paper, I introduce the alternation phenomenon (Section~\ref{sec:descriptive}) and suggest a theory-driven set of factors influencing the alternation (Section~\ref{sec:analyses}).
Then, the corpus study is presented including an appropriate statistical analysis (Section~\ref{sec:corpusstudies}).
Two experiments that correlate results from the corpus study with experimental results are then reported (Section~\ref{sec:experimental}) before I sum up the paper in Section~\ref{sec:conclusion}.

@@

Finally, it was found that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be to be part of the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modeled -- even if only using two very simple proxy variables.

@@

In step (\ref{enum:data:step3}), each of these 2,365 noun–noun combinations was queried in the target constructions (\PGCa\ and \NACa) individually in each of the first ten slices of DECOW (roughly 10 billion tokens).
In order to reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law, I used all hits for mass nouns occurring less than one hundred times, but randomly sampled one hundred hits for each mass noun that occurred one hundred or more times.
The final sample contained 6,843 sentences, which was reduced to 5,063 in the manual annotation process due to removal of noisy material, erroneous hits and uninformative cases where the measure noun was in the genitive, in which case the \NACa\ cannot be distinguished from the \PGCa.
Given the careful bootstrapping and sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target constructions.%
\footnote{In a similar fashion, the 100 most frequent measure nouns occurring with plural kind nouns were bootstrapped and queried, resulting in a sample of 871 sentences.
As stated in Section~\ref{sec:germanmeasurenps}, the \NACa\ is virtually never used with plural kind nouns, and this sample was not used except for quantifying the frequency of occurrence of the constructions (67 times \NACa\ and 794 times \PGCa).
The sample is distributed in the data package accompanying this paper, however.
}

@@

The variables \textit{Kindattraction} and \textit{Measureattraction} encode the ratio with which a given kind noun lemma or measure noun lemma occurs in the \NACb\ and the \PGCd.
They were calculated from the auxiliary samples described at the end of Section~\ref{sec:gettingdata} as a log-transformed quotient.
The higher the value, the more often the noun occurs in the \PGCa (proportionally).%
\footnote{
  It could be argued that some more advanced measure of attraction strength should be used, as is done in Collostructional (or Collexeme) Analysis \citep{GriesStefanowitsch2004}, see also \cite{Gries2015a}.
  Three main points speak against such an approach in the present case (but see \citealp[246--249]{Levshina2016} for a different approach).
  First, the attraction values will be used as regressors in a GLMM, and the values resulting from collostructional approaches, \ie\ logarithmised Fisher $p$ values, have a very unfavourable distribution in the case at hand.
  They cluster around $0$, and they include values of $-\infty$.
  Second, their main use in collostructional analyses is to \textit{sort} a list of collexemes and interpret their order.
  Their concrete numerical value might not have a solid cognitive interpretation in the given context.
  Third, I tried using collexeme strength as a regressor, and the results were unsatisfactory.
}
Additionally, \textit{Kindfreq} and \textit{Measurefreq} are the logarithm-transformed frequencies per 1,000,000 words of each lemma, extracted from the frequency lists distributed by the DECOW corpus creators on their web page.
They were added to control for basic frequency effects.

In Section~\ref{sec:analyses}, it was hypothesised that classes of measure lemmas might have different preferences for the two variants.
To capture this, class information was annotated for measure lemmas.
The classification was inspired by the list in \cite[530]{Koptjevskaja2001}, but due to the low frequencies of many of the potential classes, a very coarse classification was finally used.
With typical examples and their frequencies in the final sample, the classes are:
\textit{Physical} (abstract precisely measurable units such as \textit{Liter} `litre', \textit{Meter} `metre', \textit{Gramm} `gram'; $f=1,968$),
\textit{Container} (\textit{Eimer} `bucket'; $f=740$), 
\textit{Amount} (\textit{Menge} `amount'; $f=1,364$), 
\textit{Portion} (natural portions like \textit{Happen} `bite' or \textit{Krümel} `crumb'; $f=713$).
The few lemmas that did not fit into either of these classes were labeled \textit{Rest} ($f=278$).

@@

Finally, one variable as added as nuisance variable in the context of the present study.
It was reported in the literature that MNPs in the dative and with a masculine or neuter kind noun favour the \PGCa\ more than the corresponding nominative and accusative MNPs \citep{Hentschel1993,Zimmer2015}.
As an example, \textit{mit einem Stück frischen Brots} `with a piece of fresh bread' (\PGCa) would be preferred over \textit{mit einem Stück frischem Brot} (\NACa).
As with all the examples, native speakers of German will most likely notice that differences are subtle.
To control for this effect, the case of the measure noun was manually annotated (variable \textit{Measurecase}).

@@

In this section, I report the results of fitting a multilevel model to the data using R \citep{R}, \textit{lme4} \citep{lme4} for Maximum Likelihood Estimation, and \textit{rstanarm} \citep{rstanarm} for Markov-Chain Monte Carlo estimation (see Section~\ref{sec:rightstatistics}).
The purpose is to model the influence of the regressors specified in Table~\ref{tab:variables} on the probability that either the \PGCa\ is chosen over the \NACa.
All regressors from Table~\ref{tab:variables} were included, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The sample size was $n=5,063$ with $1,1344$ cases of \PGCa\ and $3,929$ cases of \NACa.
The results of the estimation are shown in Table~\ref{tab:bigtable} and in Figure~\ref{fig:fixeffs}.
The regressors with the measure lemma as their unit of reference have no within-measure lemma variance, and the \textit{glmer} function automatically estimates them as \textit{group level predictors} (or \textit{second-level effects}), cf.\ \cite[265--269,302--304]{GelmanHill2006}.
The same goes for those listed with the kind lemma as their unit of reference.
Given the coding of the response variable, coefficients leaning to the positive side can be interpreted as favouring the \PGCa.

@@

Standard diagnostics for the Maximum Likelihood Estimation (MLE) show that the model quality is quite high.
Generalised variance inflation factors for the regressors were calculated to check for multicollinearity \citep{FoxMonette1992,ZuurEa2010}, and none of the corrected $\text{GVIF}^{1/2\text{df}}$ was higher than $1.6$.
Nakagawa \& Schielzeth's pseudo-coefficient of determination is $R_m^2=0.409$ and $R^2_c=0.495$ (see \citealp{Gries2015} for a linguist-friendly introduction to these $R^2$ measures, or else \citealp{NakagawaSchielzeth2013}).
The rate of correct predictions is $0.843$, which means a proportional reduction of error of $\lambda=0.297$.
The measure lemma intercepts have a standard deviation of $\sigma_{\text{Measurelemma}}=0.448$, the kind lemma intercept $\sigma_{\text{Kindlemma}}=0.604$.

@@

The results reported in Section~\ref{sec:corpushierarchicalmodel} generally confirm the hypotheses from Section~\ref{sec:analyses}.
First, the prototypicality effect related to the non-alternating \PGCd\ and \NACb\ can be shown, see the effect plots in Figure~\ref{fig:eff:attraction}.%
\footnote{Effect plots were created using the \textit{effects} package \citep{Fox2003}.
They show the changes in probability for the outcome (y axis) dependent on values of a regressor (x axis), at typical values of all other regressors.}
The effect is mostly as expected:
if a lemma appears relatively more often in the \PGCd\ (compared to its frequency in the \NACb), the more often the \PGCa\ is chosen over the \NACa\ with this specific lemma.
The effect for measure nouns is stronger, and it was estimated with higher precision.

@@

The first experiment tests preferences for constructions explicitly.
\cite{FordBresnan2013} use the \textit{split-100} task in which participants have to distribute $100$ points between the alternatives, assigning more points to more natural sounding alternative.
In essence, participants distribute a probability mass between two variants, which is intended to produce more subtle results compared to a two-alternative forced-choice task \cite{Rosenbach2013}, where participants have to choose one of two variants.
The split-100 paradigm has been criticised in \cite{ArppeJaervikivi2007}.
The criticism was reiterated in \cite{DivjakEa2016}, where they use a forced-choice task.
In \cite{VerhoevenTemme2017}, it was shown that results from forced-choice and split-100 experiments mostly converge, which is expected in the long run if a probability is turned into a binary decision.
I present a forced-choice experiment in Section~\ref{sec:exp:fc} and not a split-100 experiment, mainly because in a dry-run of the experiment, participants complained about the unnaturalness of the task of distributing a probability mass across two variants.
Participants had to chose between two sentences differing only in that one contained the \NACa\ and the other contained the \PGCa.
The analysis compares the probabilities assigned to stimuli by the corpus-derived model with the frequency with which participants chose the variants for the same stimuli.

@@

\begin{table}
  \centering
  \begin{tabular}[h]{rcc}
     & Masculine\slash Neuter & Feminine \\
     \midrule
     \textbf{high prob.\ for PGC\Subsf{adj}} & 4 sentences\Supsf{a} & 4 sentences\Supsf{a} \\
     \textbf{low prob.\ for PGC\Subsf{adj}} & 4 sentences\Supsf{a} & 4 sentences\Supsf{a} \\
  \end{tabular}
  \caption{The four groups of sentences chosen as stimuli (\Supsf{a}Among the 4 sentences, combinations of important factor values were made unique whenever possible.)}
  \label{tab:experiment1:design}
\end{table}

@@

A certain amount of the variance can be accounted for by idiosyncrasies of single sentences ($\sigma_{\text{Sentence}}=1.785$, $\sigma^{\text{Modelprediction}}_{\text{Sentence}}=5.996$, $16$ levels).%
\footnote{I use $\sigma^f_r$ to denote the standard deviation of the of the random intercepts for the fixed effect $f$ varying by random effect $r$.}
Also, among participants, there are clearly different preferences ($\sigma_{\text{Participant}}=0.781$, $\sigma^{\text{Modelprediction}}_{\text{Participant}}=0.484$, $24$ levels).
On the extreme sides, one participant chose the \PGCa\ in $13$ of $16$ cases, and two participants only chose it in $5$ of $16$ cases.
The regressor \textit{Modelprediction} achieves $\mpPB=0.007$ ($1,000$ replications) and is estimated at $5.408$ relative to an intercept of $-1.304$.
The confidence interval from a parametric bootstrap ($1,000$ replications, percentile method) for the regressor is acceptable but slightly large with a lower bound of $1.626$ and an upper bound of $8.397$.
Nakagawa \& Schielzeth's pseudo-coefficients of determination are $R^2_{m}=0.227$ and $R^2_{c}=0.561$, which means that over 22\% of the variance in the data can be explained by considering only the predictions from the corpus-based GLMM.
The effect display for the single fixed regressor \textit{Modelprediction} is given in Figure \ref{fig:afc:effects}.
The result is very clear.
The higher the probability of the \PGCa\ predicted from usage-data, the more often participants chose the \PGCa\ variant in the forced-choice task.
In summary, the forced-choice experiment clearly succeeded in corroborating the results from the corpus study in as much as the preferences extracted from usage data correspond to native speakers' choices.

@@

Conretely, the exact same stimuli as in the forced-choice experiments were used.
Each participant read both the $16$ sentences with the variant predicted by the corpus model and the $16$ modified sentences with the variant that the corpus model does not predict.%
\footnote{Notice that lemmas and their frequencies as well as lemma classes are included as regressors in the corpus-based GLMM, and there was consequently no additional controlling of lemma frequencies, etc.}
To minimise repetition effects, the stimuli for each participant were separated into two blocks of $16$ targets and $33$ fillers per block.
In the experiment, participants first read all sentences from the first block, then all sentences from the second block.
It was made sure that from each target sentence pair, one sentence was assigned to the first block, and the other sentence to the other block.
The assignment of members of the individual sentence pairs to the blocks was randomised for each participant individually, and so was the order within each block.
The sentences from each pair of variants were kept as far apart as possible.
The fillers also came in pairs such that the second block exclusively contained sentences to which participants had been exposed in the first block in slightly modified form.
In total, each participant read $98$ sentences.
After each sentence, participants had to answer simple (non-metalinguistic) yes-no questions about the previous sentence as distractors.
The distractor questions were different ones in the first and the second block.
There were $38$ (different) participants recruited in exactly the same manner as for the experiment reported in Section~\ref{sec:exp:fc}.
None of them had ever participated in any kind of reading experiment, and none of them took part in the first experiment (and vice versa).
The experiment was conducted using \textit{PsychoPy}.

The reading times were residualised per speaker based on the reading times of all words (not just the targets) by that speaker.
The adjective and the kind noun (\ie\ the constituents bearing the critical case markers) were used as the target region, such as the bracketed words in the example \textit{zwei Gläser} [\textit{sprudelndes Wasser}] `two glasses of sparkling water'.
Outliers farther than 2 inter-quartile ranges from the mean logarithmised residual time were removed ($64$ data points), resulting in a total number of $n=1,152$ observations.
An LMM was specified with the logarithmised residual reading times as the response variable, and the probabilities derived from the corpus GLMM (\textit{Modelprediction}) as the main regressor.
Since the corpus GLMM predicts the probability of the \PGCa\ (vis-à-vis the \NACa), it is expected that reading times are positively correlated (longer reading times) with the probability if the stimulus actually contains the \NACa, and negatively correlated (shorter reading times) if the stimulus actually contains the \PGCa.
To account for this, an interaction between \textit{Modelprediction} and \textit{Construction} (levels \textit{PGCa} and \textit{NACa}) was added to the model.
Furthermore, the position ($1$--$98$) of the sentence in the individual experiment (\textit{Position} was included as a fixed effect to control for increasing reading speed during experiment runs.
Random intercepts were specified for \textit{Participant} and \textit{Item} (the $16$ sentence pairs are one \textit{Item} each).%
\footnote{I tried random slopes in order to keep the random effect structure maximal \citep{BarrEa2013}, but it was impossible to get the algorithm to converge due to the added complexity of the interaction.}

>>>>

In the remainder of the paper, I introduce the alternation phenomenon (Section~\ref{sec:descriptive}) and suggest a theory-driven set of factors influencing the alternation (Section~\ref{sec:analyses}).
Then, the corpus study is presented including an appropriate statistical analysis (Section~\ref{sec:corpusstudies}).
Two experiments that correlate results from the corpus study with experimental results are then reported (Section~\ref{sec:experimental}) before I sum up the paper in Section~\ref{sec:conclusion}.

@@

Finally, it was found that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be part of what defines the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modeled -- even if only using two very simple proxy variables.

@@

In step (\ref{enum:data:step3}), each of these 2,365 noun–noun combinations was queried in the target constructions (\PGCa\ and \NACa) individually in each of the first ten slices of DECOW (roughly 10 billion tokens).
In order to reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law, I used all hits for nouns with a frequency up to $100$, and a sample of $100$ of all those with higher frequency.
The final sample contained 6,843 sentences, which was reduced to 5,063 in the manual annotation process due to removal of noisy material, erroneous hits and uninformative cases where the measure noun was in the genitive, in which case the \NACa\ cannot be distinguished from the \PGCa.
Given the careful bootstrapping and sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target constructions.%
\footnote{In a similar fashion, the 100 most frequent measure nouns occurring with plural kind nouns were bootstrapped and queried, resulting in a sample of 871 sentences.
As stated in Section~\ref{sec:germanmeasurenps}, the \NACa\ is virtually never used with plural kind nouns, and this sample was not used except for quantifying the frequency of occurrence of the constructions (67 times \NACa\ and 794 times \PGCa).
The sample is distributed in the data package accompanying this paper, however.
}

@@

The variables \textit{Kindattraction} and \textit{Measureattraction} encode the ratio with which a given kind noun lemma or measure noun lemma occurs in the \PGCd\ and the  \NACb.
They were calculated from the auxiliary samples described at the end of Section~\ref{sec:gettingdata} as a log-transformed quotient.
The higher the value, the more often the noun occurs in the \PGCd (proportionally).%
\footnote{
  It could be argued that some more advanced measure of attraction strength should be used, as is done in Collostructional (or Collexeme) Analysis \citep{GriesStefanowitsch2004}, see also \cite{Gries2015a}.
  Three main points speak against such an approach in the present case (but see \citealp[246--249]{Levshina2016} for a different approach).
  First, the attraction values will be used as regressors in a GLMM, and the values resulting from collostructional approaches, \ie\ logarithmised Fisher $p$ values, have a very unfavourable distribution in the case at hand.
  They cluster around $0$, and they include values of $-\infty$.
  Second, their main use in collostructional analyses is to \textit{sort} a list of collexemes and interpret their order.
  Their concrete numerical value might not have a solid cognitive interpretation in the given context.
  Third, I tried using collexeme strength as a regressor, and the results were unsatisfactory.
}
Additionally, \textit{Kindfreq} and \textit{Measurefreq} are the logarithm-transformed frequencies per 1,000,000 words of each lemma, extracted from the frequency lists distributed by the DECOW corpus creators on their web page.
They were added to control for basic frequency effects.

In Section~\ref{sec:analyses}, it was hypothesised that classes of measure lemmas might have different preferences for the two variants.
To capture this, class information was annotated for measure lemmas.
The classification was inspired by the list in \citet[530]{Koptjevskaja2001}, but due to the low frequencies of many of the potential classes, a very coarse classification was finally used.
With typical examples and their frequencies in the final sample, the classes are:
\textit{Physical} (abstract precisely measurable units such as \textit{Liter} `litre', \textit{Meter} `metre', \textit{Gramm} `gram'; $f=1,968$),
\textit{Container} (\textit{Eimer} `bucket'; $f=740$), 
\textit{Amount} (\textit{Menge} `amount'; $f=1,364$), 
\textit{Portion} (natural portions like \textit{Happen} `bite' or \textit{Krümel} `crumb'; $f=713$).
The few lemmas that did not fit into either of these classes were labeled \textit{Rest} ($f=278$).

@@

Finally, one variable was added as nuisance variable in the context of the present study.
It was reported in the literature that MNPs in the dative and with a masculine or neuter kind noun favour the \PGCa\ more than the corresponding nominative and accusative MNPs \citep{Hentschel1993,Zimmer2015}.
As an example, \textit{mit einem Stück frischen Brots} `with a piece of fresh bread' (\PGCa) would be preferred over \textit{mit einem Stück frischem Brot} (\NACa).
As with all the examples, native speakers of German will most likely notice that differences are subtle.
To control for this effect, the case of the measure noun was manually annotated (variable \textit{Measurecase}).

In this section, I motivate the choice of statistical models which I use in Section~\ref{sec:corpushierarchicalmodel}.
Readers might think that the method used here for modeling grammatical alternations -- namely (Hierarchical) Logistic Regression\slash Generalised Linear (Mixed) Models or GLM(M)s -- does not require much motivation.
After all, GLMMs have been established as the major tool in the analysis of alternation phenomena.
All studies mentioned at the outset of Section~\ref{sec:cogocl} use some form of regression\slash GLM(M).
Over the past few years, however, modified or alternative methods have been proposed.
From among these methods, I just make a few remarks on Bayesian estimation (see \citealp{GelmanEa2014}), as it was proposed in \cite{Levshina2016,Divjak2016a}, for example.
Conceptually, I see three points of discussion that should be kept apart.
First, Bayesian methods are sometimes touted as superior tools for scientific inference compared to frequentist methods.
Second, it has been proposed that the Bayesian interpretation of probability is more cognitively adequate for the modeling of linguistic data.
Third, and very specific to this paper, given established methods in the modeling of alternation and variation, it has to be decided whether so-called Bayesian methods lead to substantially different results.

@@

In this section, I report the results of fitting a multilevel model to the data using R \citep{R}, \textit{lme4} \citep{lme4} for Maximum Likelihood Estimation, and \textit{rstanarm} \citep{rstanarm} for Markov-Chain Monte Carlo estimation (see Section~\ref{sec:rightstatistics}).
The purpose is to model the influence of the regressors specified in Table~\ref{tab:variables} on the probability that the \PGCa\ is chosen over the \NACa.
All regressors from Table~\ref{tab:variables} were included, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The sample size was $n=5,063$ with $1,1344$ cases of \PGCa\ and $3,929$ cases of \NACa.
The results of the estimation are shown in Table~\ref{tab:bigtable} and in Figure~\ref{fig:fixeffs}.
The regressors with the measure lemma as their unit of reference have no within-measure lemma variance, and the \textit{glmer} function automatically estimates them as \textit{group level predictors} (or \textit{second-level effects}), cf.\ \citet[265--269,302--304]{GelmanHill2006}.
The same goes for those listed with the kind lemma as their unit of reference.
Given the coding of the response variable, coefficients leaning to the positive side can be interpreted as favouring the \PGCa.

@@

Standard diagnostics for the Maximum Likelihood Estimation (MLE) show that the model quality is quite high.
Generalised variance inflation factors for the regressors were calculated to check for multicollinearity \citep{FoxMonette1992,ZuurEa2010}, and none of the corrected $\text{GVIF}^{1/2\text{df}}$ was higher than $1.6$.
Nakagawa \& Schielzeth's pseudo-coefficient of determination is $R_m^2=0.409$ and $R^2_c=0.495$ (see \citealp{Gries2015} for a basic introduction to these $R^2$ measures, or else \citealp{NakagawaSchielzeth2013}).
The rate of correct predictions is $0.843$, which means a proportional reduction of error of $\lambda=0.297$.
The lemma intercepts have standard deviations of $\sigma_{\text{Measurelemma}}=0.448$ and $\sigma_{\text{Kindlemma}}=0.604$.

@@

The results reported in Section~\ref{sec:corpushierarchicalmodel} generally confirm the hypotheses from Section~\ref{sec:analyses}.
First, the prototypicality effect related to the non-alternating \PGCd\ and \NACb\ can be shown, see the effect plots in Figure~\ref{fig:eff:attraction}.%
\footnote{Effect plots were created using the \textit{effects} package \citep{Fox2003}.
They show the changes in probability for the outcome (y axis) dependent on values of a regressor (x axis), at typical values of all other regressors.
The vertical bars (categorical variables) and grey areas (continuous variables) are asymptotic $95\%$ confidence intervals calculated from \textit{glmer}.
They are not bootstrapped.}
The effect is mostly as expected:
if a lemma appears relatively more often in the \PGCd\ (compared to its frequency in the \NACb), the more often the \PGCa\ is chosen over the \NACa\ with this specific lemma.
The effect for measure nouns is stronger, and it was estimated with higher precision.

@@

The first experiment tests preferences for constructions explicitly.
\cite{FordBresnan2013} use the \textit{split-100} task in which participants have to distribute $100$ points between the alternatives, assigning more points to more natural sounding alternative.
In essence, participants distribute a probability mass between two variants, which is intended to produce more subtle results compared to a two-alternative forced-choice task \citep{Rosenbach2013}, where participants have to choose one of two variants.
The split-100 paradigm has been criticised in \cite{ArppeJaervikivi2007}.
The criticism was reiterated in \cite{DivjakEa2016}, where they use a forced-choice task.
In \cite{VerhoevenTemme2017}, it was shown that results from forced-choice and split-100 experiments mostly converge, which is expected in the long run if a probability is turned into a binary decision.
I present a forced-choice experiment in Section~\ref{sec:exp:fc} and not a split-100 experiment, mainly because in a dry-run of the experiment, participants complained about the unnaturalness of the task of distributing a probability mass across two variants.
Participants had to chose between two sentences differing only in that one contained the \NACa\ and the other contained the \PGCa.
The analysis compares the probabilities assigned to stimuli by the corpus-derived model with the frequency with which participants chose the variants for the same stimuli.

@@

\begin{table}
  \centering
  \begin{tabular}[h]{rcc}
     & Masculine\slash Neuter & Feminine \\
     \midrule
     \textbf{high prob.\ for PGC\Subsf{adj}} & 4 sentences & 4 sentences \\
     \textbf{low prob.\ for PGC\Subsf{adj}} & 4 sentences & 4 sentences \\
  \end{tabular}
  \caption{The four groups of sentences chosen as stimuli; in each group of four sentences, combinations of important factor values were made unique whenever possible}
  \label{tab:experiment1:design}
\end{table}

@@

A certain amount of the variance can be accounted for by idiosyncrasies of single sentences ($\sigma_{\text{Sentence}}=1.785$, $\sigma^{\text{Modelprediction}}_{\text{Sentence}}=5.996$, $16$ levels).%
\footnote{I use $\sigma^f_r$ to denote the standard deviation of the random intercepts for the fixed effect $f$ varying by random effect $r$.}
Also, among participants, there are clearly different preferences ($\sigma_{\text{Participant}}=0.781$, $\sigma^{\text{Modelprediction}}_{\text{Participant}}=0.484$, $24$ levels).
On the extreme sides, one participant chose the \PGCa\ in $13$ of $16$ cases, and two participants only chose it in $5$ of $16$ cases.
The regressor \textit{Modelprediction} achieves $\mpPB=0.007$ ($1,000$ replications) and is estimated at $5.408$ relative to an intercept of $-1.304$.
The confidence interval from a parametric bootstrap ($1,000$ replications, percentile method) for the regressor is acceptable but slightly large with a lower bound of $1.626$ and an upper bound of $8.397$.
Nakagawa \& Schielzeth's pseudo-coefficients of determination are $R^2_{m}=0.227$ and $R^2_{c}=0.561$, which means that over 22\% of the variance in the data can be explained by considering only the predictions from the corpus-based GLMM.
The effect display for the single fixed regressor \textit{Modelprediction} is given in Figure \ref{fig:afc:effects}.
The result is very clear.
The higher the probability of the \PGCa\ predicted from usage-data, the more often participants chose the \PGCa\ variant in the forced-choice task.
In summary, the forced-choice experiment clearly succeeded in corroborating the results from the corpus study in as much as the preferences extracted from usage data correspond to native speakers' choices.

@@

Conretely, the exact same stimuli as in the forced-choice experiments were used.
Each participant read both the $16$ sentences with the variant predicted by the corpus model and the $16$ modified sentences with the variant that the corpus model does not predict.%
\footnote{Notice that lemmas and their frequencies as well as lemma classes are included as regressors in the corpus-based GLMM, and there was consequently no additional controlling of lemma frequencies, etc.}
To minimise repetition effects, the stimuli for each participant were separated into two blocks of $16$ targets and $33$ fillers per block.
In the experiment, participants first read all sentences from the first block, then all sentences from the second block.
It was made sure that from each target sentence pair, one sentence was assigned to the first block, and the other sentence to the other block.
The assignment of members of the individual sentence pairs to the blocks was randomised for each participant individually, and so was the order within each block.
The sentences from each pair of variants were kept as far apart as possible.
The fillers also came in pairs such that the second block exclusively contained sentences to which participants had been exposed in the first block in slightly modified form.
In total, each participant read $98$ sentences.
After each sentence, participants had to answer simple (non-metalinguistic) yes-no questions about the previous sentence as distractors.
The distractor questions were different ones in the first and the second block.
There were $38$ (different) participants recruited in exactly the same manner as for the experiment reported in Section~\ref{sec:exp:fc}.
None of them had ever participated in any kind of reading experiment, and none of them took part in the first experiment.
The experiment was conducted using \textit{PsychoPy}.

The reading times were residualised per speaker based on the reading times of all words (not just the targets) by that speaker.
The adjective and the kind noun (\ie\ the constituents bearing the critical case markers) were used as the target region, such as the bracketed words in the example \textit{zwei Gläser} [\textit{spru\-deln\-des Wasser}] `two glasses of sparkling water'.
Outliers farther than 2 inter-quartile ranges from the mean logarithmised residual time were removed ($64$ data points), resulting in a total number of $n=1,152$ observations.
An LMM was specified with the logarithmised residual reading times as the response variable, and the probabilities derived from the corpus GLMM (\textit{Modelprediction}) as the main regressor.
Since the corpus GLMM predicts the probability of the \PGCa\ (vis-à-vis the \NACa), it is expected that reading times are positively correlated (longer reading times) with the probability if the stimulus actually contains the \NACa, and negatively correlated (shorter reading times) if the stimulus actually contains the \PGCa.
To account for this, an interaction between \textit{Modelprediction} and \textit{Construction} (levels \textit{PGCa} and \textit{NACa}) was added to the model.
Furthermore, the position ($1$--$98$) of the sentence in the individual experiment (\textit{Position} was included as a fixed effect to control for increasing reading speed during experiment runs.
Random intercepts were specified for \textit{Participant} and \textit{Item} (the $16$ sentence pairs are one \textit{Item} each).%
\footnote{I tried random slopes in order to keep the random effect structure maximal \citep{BarrEa2013}, but it was impossible to get the algorithm to converge due to the added complexity of the interaction.}
